---
title: "Pre-filtering VCF files and Differentiation of Monoclonal and Polyclonal samples"
author: "Paulo Manrique"
date: "2023-04-27"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Background

After the variant call process the **VCF** file generated contains some artefacts that affect downstream analysis. Alignment artifacts can occur whenever there is sufficient sequence similarity between two or more regions in the genome to confuse the alignment algorithm. This can occur when the aligner for whatever reason overestimate how uniquely a read maps, thereby assigning it too high of a mapping quality. It can also occur through no fault of the aligner due to gaps in the reference, which can also hide the true position to which a read should map. Even thought **GATK** solve most of these artifacts some still persists. In this tutorial we will filter out artifacts from our variant calls and generate a *'cleaner'* set of loci for downstream analysis.

# Methods

## Section 1: Filtering of VCF files

### Setp 1: Filter of coding regions.

In this step the `.vcf` file is still too large to be manipulated in our personal computer, for that reason this step will be run in the server.

For this first step we require the following files:

-   `.vcf.gz` or `.vcf` files: If a `.vcf.gz` is used, a `.vcf.tbi` will be also required. For this tutorial the `.vfc.gz` file that we are going to work with is: `VENPER2019_2022Partial.JointCall.filtered.combined.snpeff.vcf`.

-   `.gff`: the gff file of the reference genome will be also required in order to filter coding regions.

-   `noncore_regions.bed`: This file will be used to remove non-core regions that are difficult to sequence from the analysis. This file won't be included as it remove genes of interest located in subtelomeric regions such as duffy binding protein 2.

-   `load_libraries.R`: This R script file load all the required packages that we need to perform the analysis.

-   `functions.R`: This R script file load all the functions that we have created to perform the analysis. Within those functions we a function named `run_vcftools` that allow us to manipulate the `vcftools` package from the R environment. We have also created other functions to get information from the `.vcf` file by loading the whole file in R environment.

-   `Coding_regions_filter.R`: This R script file contains the instructions to filter the coding regions in R. This script requires 9 arguments:

    -   `wd`: Working directory where all temporal files and final outputs will be stored.

    -   `fd`: Directory where the pre-required files are stored.

    -   `gzvcf`: Path to the `.vcf.gz` file.

    -   `o`: prefix used for the temporal and final outputs.

    -   `rkeep`: regular expression to identify samples of interest.

    -   `ebed`: name of the `noncore_regions.bed` file. This file should be stored in the `fd` path.

    -   `gff`: name of the `.gff` file. This file should be stored in the `fd` path.

- `Split_chromosomes.R`: Scrip with instruction to split the large vcf file by chromosomes, and convert each of the vcf files to the rGenome fromat (a format that is easier to manipulate in r for further analysis). In addition to the previous arguments, this script contains two more arguments to speed up the analysis in the server, and to define the minimum read depth to call an allele.

    -   `n`: number of iterations. In the final step of this script we are going to transform our data from VCF format to rGenome format. This step is computing exhausting and consumes RAM memory. For that reason we subdivide this step in `n` instances to be able to run in the server.

    -   `thres`: Minimum read depth to call an allele in a sample.

-   `r_options_CR_filter_Pviv.sh`: bash script with the instructions to run `Coding_regions_filter.R`. This file should be stored in `wd`. The scrip should looks as follow:
```{bash, eval = FALSE}
#!/bin/bash
source /broad/software/scripts/useuse
use R-4.1
Rscript /gsap/garage-protistvector/PanSAmer/Coding_regions_filter.R \
  -wd /gsap/garage-protistvector/PanSAmer/PerVenPviv2023 \
  -fd /gsap/garage-protistvector/ColombiaData/Pviv/AnalysisTools \
  -gzvcf /seq/plasmodium/PanSAmer/JointCall/VENPER2019_2022.JointCall.filtered.combined.snpeff.vcf \
  -o PerVen_Pviv_2023 \
  -gff genes.gff \
  -n 500 \r_o   pv    
  -thres 1
```

- `uger_options_pviv.sh`: bash script with the instruction to run UGER and the bash script `r_options_CR_filter_Pviv.sh`. This file should be stored in `wd`. The scrip should looks as follow:
```{bash, eval = FALSE}
#!/bin/bash
qsub -l h_vmem=32G \
   -l h_rt=10:00:00 \
   -o /gsap/garage-protistvector/PanSAmer/output \
   /gsap/garage-protistvector/PanSAmer/r_options_CR_filter_Pviv.sh
```

- `r_options_chr_filter_Pviv.sh`: : bash script with the instructions to run `Split_chromosomes.R`. This file should be stored in `wd`. The scrip should looks as follow:
```{bash, eval = FALSE}
#!/bin/bash
source /broad/software/scripts/useuse
use R-4.1
Rscript /gsap/garage-protistvector/PanSAmer/Split_chromosomes.R \
  -wd /gsap/garage-protistvector/PanSAmer/PerVenPviv2023 \
  -fd /gsap/garage-protistvector/ColombiaData/Pviv/AnalysisTools \
  -vcf /gsap/garage-protistvector/PanSAmer/PerVenPviv2023/PerVen_Pviv_2023_CodingRegionsOnly.recode.vcf \
  -chr $SGE_TASK_ID \
  -o PerVen_Pviv_2023_thres1_CodingRegionsOnly \
  -n 500 \
  -thres 1 \
  -p Pv
```

- `uger_options_chr_pviv.sh`: bash script with the instruction to run UGER and the bash script `r_options_chr_filter_Pviv.sh`. This file should be stored in `wd`. The scrip should looks as follow:
```{bash, eval = FALSE}
#!/bin/bash
qsub -l h_vmem=32G \
   -l h_rt=10:00:00 \
   -o /gsap/garage-protistvector/PanSAmer/output2/ \
   -t 1-16 \
   /gsap/garage-protistvector/PanSAmer/r_options_chr_filter_Pviv.sh
```
As a final output this step will generate and `.RData` file containing the data in `rGenome` format.

### Step 2: Low support (Read depth) and missing data

1.1 Load the required libraries and functions:

```{r 'Loading required libraries and functions'}
setwd('~/Documents/Github/Plasmodium_WGS_analysis/Pre_filtering/')

load('backup.RData')

source('../functions_libraries/load_libraries.R')
source('../functions_libraries/functions.R')
#sourceCpp('../functions_libraries/Rcpp_functions.cpp')
```

1.2 Load the rGenome objects:
```{r 'Load the rGenome objects', eval = FALSE}
for(file in list.files('PerVen/Pviv/', pattern = '.+thres1.+.RData')){
  load(file.path('PerVen/Pviv/', file))
  rm(list = ls()[!grepl('PvP01_(\\d+|MIT|API)_v1_rGenome_object', ls())])
}

rGenome_objects = list(PvP01_01_v1_rGenome_object,
     PvP01_02_v1_rGenome_object,
     PvP01_03_v1_rGenome_object,
     PvP01_04_v1_rGenome_object,
     PvP01_05_v1_rGenome_object,
     PvP01_06_v1_rGenome_object,
     PvP01_07_v1_rGenome_object,
     PvP01_08_v1_rGenome_object,
     PvP01_09_v1_rGenome_object,
     PvP01_10_v1_rGenome_object,
     PvP01_11_v1_rGenome_object,
     PvP01_12_v1_rGenome_object,
     PvP01_13_v1_rGenome_object#,
     #'PvP01_14_v1_rGenome_object,
     #'PvP01_MIT_v1_rGenome_object,
     #'PvP01_API_v1_rGenome_object
     )

rm(list = c('PvP01_01_v1_rGenome_object',
     'PvP01_02_v1_rGenome_object',
     'PvP01_03_v1_rGenome_object',
     'PvP01_04_v1_rGenome_object',
     'PvP01_05_v1_rGenome_object',
     'PvP01_06_v1_rGenome_object',
     'PvP01_07_v1_rGenome_object',
     'PvP01_08_v1_rGenome_object',
     'PvP01_09_v1_rGenome_object',
     'PvP01_10_v1_rGenome_object',
     'PvP01_11_v1_rGenome_object',
     'PvP01_12_v1_rGenome_object',
     'PvP01_13_v1_rGenome_object'#,
     #'PvP01_14_v1_rGenome_object',
     #'PvP01_MIT_v1_rGenome_object',
     #'PvP01_API_v1_rGenome_object'
     ))

source('../functions_libraries/functions.R')
Pviv_rGenome_object = rGenome(gt = NULL,
                              loci_table = NULL,
                              metadata = NULL)

metadata = rGenome_objects[[1]]@metadata

for(obj in 1:length(rGenome_objects)){
  Pviv_rGenome_object@gt = rbind(Pviv_rGenome_object@gt, rGenome_objects[[obj]]@gt)
  Pviv_rGenome_object@loci_table = rbind(Pviv_rGenome_object@loci_table, rGenome_objects[[obj]]@loci_table)
  rGenome_objects[[obj]] = NULL
  
}

Pviv_rGenome_object@metadata = rbind(Pviv_rGenome_object@metadata, rGenome_objects[[obj]]@metadata)

rm(list = c('rGenome_objects', 'obj'))

# save.image('Prefiltered_Pviv_1st.RData')
# load('Prefiltered_Pviv_1st.RData')
```

2. Add Metadata.

```{r 'Codigo de R', eval = FALSE}

# There are some spelling mistakes in the sample codes of the sequencing files, The table VENPER2019_2022.csv match the incorrect codes with the correct codes
external_metadata = read.csv('../GeneticDiversity/merged_metadata.csv')
PerVen_seq_codes = read.csv('PerVen/VENPER2019_2022.csv')

# Remove Duplicated records

View(PerVen_seq_codes[PerVen_seq_codes$PreferedSampleID %in% PerVen_seq_codes$PreferedSampleID[duplicated(PerVen_seq_codes$PreferedSampleID)],])



external_metadata = external_metadata[!duplicated(external_metadata$Sample_id),]

PerVen_seq_codes = PerVen_seq_codes[!duplicated(PerVen_seq_codes$PreferedSampleID),]

# merge incorrect and correct codes with metadata
external_metadata = merge(external_metadata, PerVen_seq_codes, by.x = 'Sample_id', by.y = 'AlternateSampleID', all.x = T)

external_metadata %<>% mutate(PreferedSampleID = case_when(
  is.na(PreferedSampleID) ~ Sample_id,
  !is.na(PreferedSampleID) ~ PreferedSampleID
))

# Add metadata to the rGenome object
Pviv_rGenome_object@metadata = merge(Pviv_rGenome_object@metadata, external_metadata[,c("PreferedSampleID", "Sample_id", "Study", "Country", 'Date_of_Collection', "Year_of_Collection",'Subnational_level0', 'Subnational_level1', "Subnational_level2")], by.x = 'Sample_id', by.y = 'PreferedSampleID', all.x = T)

# Sorting samples
rownames(Pviv_rGenome_object@metadata) = Pviv_rGenome_object@metadata$Sample_id

Pviv_rGenome_object@metadata = Pviv_rGenome_object@metadata[colnames(Pviv_rGenome_object@gt),]

Pviv_rGenome_object@metadata[is.na(Pviv_rGenome_object@metadata$Country),][['Country']] = 'Venezuela'

```

#### Setp 2.1: Site coverage (Read depth by site)

3. Calculate the total (sum over all alleles and all samples) and mean (arithmetic mean of the total read depth of the samples) read depth by site using the function `Summarise_ReadDepth()`.
```{r, eval = FALSE}
# Calculate the total and mean read depth per each site per each country using the function Summarise_ReadDepth.

Read_Depth_Summ = NULL

for(w in 1:100){
  start = Sys.time()
  Read_Depth_Summ = rbind(Read_Depth_Summ, Summarise_ReadDepth(obj = Pviv_rGenome_object, by = 'Country', w = w, n = 100))
  end = Sys.time()
  print(paste0('Iteration ',w , ' took ', end - start, ' secs'))
}

# Transform the Read_Depth_Summ table from wide to long format

## Extract data from Peru
Read_Depth_Summ_Peru = data.frame(Chrom = 
                                  gsub('_\\d+$','',
                                       rownames(Read_Depth_Summ)),
                                  Pos = gsub('^.+_','',
                                             rownames(Read_Depth_Summ)),
                                  Country = 'Peru',
                                  Read_Depth_Summ[grepl('_Per',
                                                   colnames(Read_Depth_Summ))])

colnames(Read_Depth_Summ_Peru) = gsub('_Per.$','',
                                      colnames(Read_Depth_Summ_Peru))
## Extract data Venezuela
Read_Depth_Summ_Venezuela = data.frame(Chrom = 
                                         gsub('_\\d+$','',
                                                    rownames(Read_Depth_Summ)),
                                         Pos = gsub('^.+_','',
                                                    rownames(Read_Depth_Summ)),
                                         Country = 'Venezuela',
                                         Read_Depth_Summ[grepl('_Ven',
                                                   colnames(Read_Depth_Summ))])

colnames(Read_Depth_Summ_Venezuela) = gsub('_Ven.+$','',
                                           colnames(Read_Depth_Summ_Venezuela))

## Extract data from the total Population
Read_Depth_Summ_Total = data.frame(Chrom = gsub('_\\d+$','',
                                                rownames(Read_Depth_Summ)),
                             Pos = gsub('^.+_','',
                                        rownames(Read_Depth_Summ)),
                             Country = 'Total',
                             Read_Depth_Summ[grepl('_Total$',
                                                   colnames(Read_Depth_Summ))])

colnames(Read_Depth_Summ_Total) = gsub('_Total$','',
                                       colnames(Read_Depth_Summ_Total))


Read_Depth_Summ = rbind(Read_Depth_Summ_Peru,
                        Read_Depth_Summ_Venezuela,
                        Read_Depth_Summ_Total)

rm(list = c('Read_Depth_Summ_Peru',
            'Read_Depth_Summ_Venezuela',
            'Read_Depth_Summ_Total'))

Read_Depth_Summ$Pos = as.integer(Read_Depth_Summ$Pos)
Read_Depth_Summ$Country = factor(Read_Depth_Summ$Country,
                                 levels = c('Peru',
                                            'Venezuela',
                                            'Total'))
```

4. Plot the total read depth by site, by country, and by chromosome.
```{r, fig.height = 16, fig.width = 8, fig.cap= '**Figure 1:** Total read depth by site.'}
Read_Depth_Summ %>% ggplot(aes(x = Pos, y = Total_ReadDepth, color = Country, group = Country)) +
  geom_line() +
  facet_grid(Chrom~Country, scales = 'free_y') +
  scale_color_manual(values = c('dodgerblue3', 'firebrick3', 'black')) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

5. Plot the mean read depth by site, by country, and by chromosome.
```{r, fig.height = 16, fig.width = 8, fig.cap= '**Figure 2:** Mean read depth by site.'}
Read_Depth_Summ %>% ggplot(aes(x = Pos, y = mean_ReadDepth, color = Country, group = Country)) +
  geom_line() +
  facet_grid(Chrom~Country, scales = 'free_y') +
  scale_color_manual(values = c('dodgerblue3', 'firebrick3', 'black')) +
  theme_bw() +
  theme(axis.text.x = element_blank())
```

#### Step 2.2: Remove samples with low amplification rate.

6. Calculate the proportion of amplified loci (amplification rate), with a coverage equals or greater than 1, 2, 3, 4 and 5, of the samples.

```{r, eval = FALSE}
# Sample amplification rate including all detected reads
SampAmpRate_thres0 = SampleAmplRate(Pviv_rGenome_object, update = FALSE, threshold = NULL)
SampAmpRate_thres0 = data.frame(Sample_id = names(SampAmpRate_thres0),
                                AmpRate = SampAmpRate_thres0,
                                Threshold = 1)
SampAmpRate_thres0 = merge(SampAmpRate_thres0,
                           Pviv_rGenome_object@metadata[,c('Sample_id', 'Country')],
                           by = 'Sample_id',
                           all.x = TRUE)

# Sample amplification rate including only sites supported by at least 2 reads
SampAmpRate_thres1 = SampleAmplRate(Pviv_rGenome_object, update = FALSE, threshold = 1, n = 100)
SampAmpRate_thres1 = data.frame(Sample_id = names(SampAmpRate_thres1),
                                AmpRate = SampAmpRate_thres1,
                                Threshold = 2)
SampAmpRate_thres1 = merge(SampAmpRate_thres1,
                           Pviv_rGenome_object@metadata[,c('Sample_id', 'Country')],
                           by = 'Sample_id',
                           all.x = TRUE)

# Sample amplification rate including only sites supported by at least 3 reads
SampAmpRate_thres2 = SampleAmplRate(Pviv_rGenome_object, update = FALSE, threshold = 2, n = 100)
SampAmpRate_thres2 = data.frame(Sample_id = names(SampAmpRate_thres2),
                                AmpRate = SampAmpRate_thres2,
                                Threshold = 3)
SampAmpRate_thres2 = merge(SampAmpRate_thres2,
                           Pviv_rGenome_object@metadata[,c('Sample_id', 'Country')],
                           by = 'Sample_id',
                           all.x = TRUE)

# Sample amplification rate including only sites supported by at least 4 reads
SampAmpRate_thres3 = SampleAmplRate(Pviv_rGenome_object, update = FALSE, threshold = 3, n = 100)
SampAmpRate_thres3 = data.frame(Sample_id = names(SampAmpRate_thres3),
                                AmpRate = SampAmpRate_thres3,
                                Threshold = 4)
SampAmpRate_thres3 = merge(SampAmpRate_thres3,
                           Pviv_rGenome_object@metadata[,c('Sample_id', 'Country')],
                           by = 'Sample_id',
                           all.x = TRUE)

# Sample amplification rate including only sites supported by at least 5 reads
SampAmpRate_thres4 = SampleAmplRate(Pviv_rGenome_object, update = FALSE, threshold = 4, n = 100)
SampAmpRate_thres4 = data.frame(Sample_id = names(SampAmpRate_thres4),
                                AmpRate = SampAmpRate_thres4,
                                Threshold = 5)
SampAmpRate_thres4 = merge(SampAmpRate_thres4,
                           Pviv_rGenome_object@metadata[,c('Sample_id', 'Country')],
                           by = 'Sample_id',
                           all.x = TRUE)

# Combining all vectors of SampAmpRate
SampAmpRate_summ = rbind(SampAmpRate_thres0,
                    SampAmpRate_thres1,
                    SampAmpRate_thres2,
                    SampAmpRate_thres3,
                    SampAmpRate_thres4)

rm(list = c('SampAmpRate_thres0',
            'SampAmpRate_thres1',
            'SampAmpRate_thres2',
            'SampAmpRate_thres3',
            'SampAmpRate_thres4'))
```

7. Plot the proportion of samples that success to amplify with an specific amplification rate.
```{r, fig.height=4, fig.width=8, fig.cap='Decay curve of the percentage of passed samples as a function of their amplification rate. Line color assigned by min read depth (coverage) to call an allele'}
SampAmpRate_summ %>% group_by(Country, Threshold) %>%
  summarise(AmpRate5 = round(100*sum(AmpRate >= .05)/n(), 1),
            AmpRate10 = round(100*sum(AmpRate >= .10)/n(), 1),
            AmpRate15 = round(100*sum(AmpRate >= .15)/n(), 1),
            AmpRate20 = round(100*sum(AmpRate >= .20)/n(), 1),
            AmpRate25 = round(100*sum(AmpRate >= .25)/n(), 1),
            AmpRate30 = round(100*sum(AmpRate >= .30)/n(), 1),
            AmpRate35 = round(100*sum(AmpRate >= .35)/n(), 1),
            AmpRate40 = round(100*sum(AmpRate >= .40)/n(), 1),
            AmpRate45 = round(100*sum(AmpRate >= .45)/n(), 1),
            AmpRate50 = round(100*sum(AmpRate >= .50)/n(), 1),
            AmpRate55 = round(100*sum(AmpRate >= .55)/n(), 1),
            AmpRate60 = round(100*sum(AmpRate >= .60)/n(), 1),
            AmpRate65 = round(100*sum(AmpRate >= .65)/n(), 1),
            AmpRate70 = round(100*sum(AmpRate >= .70)/n(), 1),
            AmpRate75 = round(100*sum(AmpRate >= .75)/n(), 1),
            AmpRate80 = round(100*sum(AmpRate >= .80)/n(), 1),
            AmpRate85 = round(100*sum(AmpRate >= .85)/n(), 1),
            AmpRate90 = round(100*sum(AmpRate >= .90)/n(), 1),
            AmpRate95 = round(100*sum(AmpRate >= .95)/n(), 1),
            AmpRate100 = round(100*sum(AmpRate >= 1)/n(), 1)) %>%
  pivot_longer(cols = paste0('AmpRate', seq(5, 100, 5)),
               values_to = 'Percentage',
               names_to = 'AmpRate') %>%
  mutate(AmpRate = as.numeric(gsub('AmpRate','', AmpRate)))%>%
  ggplot(aes(x = AmpRate, y = Percentage, color = as.factor(Threshold), group = as.factor(Threshold))) +
  geom_line() +
  geom_vline(xintercept = 75, linetype = 2) +
  facet_grid(.~Country) +
  theme_bw() +
  labs(x = '% of amplified loci (amplification rate)', y = '% of Samples', color = 'Min Coverage')
```
8. Table of the number of samples and the proportion of samples that amplify more than 50% of their loci with a coverage equals or greater than 5
```{r}
SampAmpRate_summ %>% group_by(Country, Threshold) %>%
  summarise(nSample_w_AmpRate50 = sum(AmpRate >= .5),
            nSamples = n(),
            prop_of_samples_w_AmpRate50 = round(100*sum(AmpRate >= .5)/n(), 1))
```

9. Calculate the proportion of amplified samples per each loci using a coverage of 1 and 5 to call an allele.
```{r, eval = FALSE}
# Create an rGenome object per  each country

Peru_rGenome = filter_samples(obj = Pviv_rGenome_object,
                             v = Pviv_rGenome_object@metadata$Country == 'Perú')


Venezuela_rGenome = filter_samples(obj = Pviv_rGenome_object,
                             v = Pviv_rGenome_object@metadata$Country == 'Venezuela')


Peru_LocusAmpRate_thres0 = LocusAmplRate(Peru_rGenome, update = FALSE, threshold = NULL)

Peru_LocusAmpRate_thres0 = 
  data.frame(Locus_id = names(Peru_LocusAmpRate_thres0),
             Chrom = gsub(
               '_\\d+$',
               '',
               names(Peru_LocusAmpRate_thres0)),
             Pos = as.integer(gsub(
               '^.+_',
               '',
               names(Peru_LocusAmpRate_thres0))),
             AmpRate = Peru_LocusAmpRate_thres0,
             Threshold = 1)

Venezuela_LocusAmpRate_thres0 = LocusAmplRate(Venezuela_rGenome, update = FALSE, threshold = NULL)

Venezuela_LocusAmpRate_thres0 = 
  data.frame(Locus_id = names(Venezuela_LocusAmpRate_thres0),
             Chrom = gsub(
               '_\\d+$',
               '',
               names(Venezuela_LocusAmpRate_thres0)),
             Pos = as.integer(gsub(
               '^.+_',
               '',
               names(Venezuela_LocusAmpRate_thres0))),
             AmpRate = Venezuela_LocusAmpRate_thres0,
             Threshold = 1)


Peru_LocusAmpRate_thres4 = LocusAmplRate(Peru_rGenome, update = FALSE, threshold = 4, n =100)

Peru_LocusAmpRate_thres4 = 
  data.frame(Locus_id = names(Peru_LocusAmpRate_thres4),
             Chrom = gsub(
               '_\\d+$',
               '',
               names(Peru_LocusAmpRate_thres4)),
             Pos = as.integer(gsub(
               '^.+_',
               '',
               names(Peru_LocusAmpRate_thres4))),
             AmpRate = Peru_LocusAmpRate_thres4,
             Threshold = 5)


Venezuela_LocusAmpRate_thres4 = LocusAmplRate(Venezuela_rGenome, update = FALSE, threshold = 4, n =100)

Venezuela_LocusAmpRate_thres4 = 
  data.frame(Locus_id = names(Venezuela_LocusAmpRate_thres4),
             Chrom = gsub(
               '_\\d+$',
               '',
               names(Venezuela_LocusAmpRate_thres4)),
             Pos = as.integer(gsub(
               '^.+_',
               '',
               names(Venezuela_LocusAmpRate_thres4))),
             AmpRate = Venezuela_LocusAmpRate_thres4,
             Threshold = 5)
```

10. Manhattan plot the of the proportion of amplified samples by loci and by country.

```{r, fig.height=8, fig.width=6, fig.cap='Manhattan plot the of the proportion of amplified samples by loci and by country'}
ggdraw() + 
  draw_plot(
    Read_Depth_Summ %>%
  filter(grepl('_01_', Chrom), Country == 'Venezuela')%>% ggplot(aes(x = Pos, y = mean_ReadDepth)) + 
  geom_line() +
  facet_grid(.~Chrom)+
  theme_bw()+
  theme(axis.text.x = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank())+
    labs(y = 'Mean read depth)'),
  x = .5,
  y = .75,
  heigh = .25,
  width = .5
  )+
  draw_plot(
    Venezuela_LocusAmpRate_thres0 %>%
  filter(grepl('_01_', Chrom))%>% ggplot(aes(x = Pos, y = AmpRate)) + 
  geom_point(alpha = .3, size = .25) +
  theme_bw()+
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank())+
    labs(y = 'Amp. rate (thres = 1)'),
  x = .5,
  y = .5,
  heigh = .25,
  width = .5
  )+
  draw_plot(
    Venezuela_LocusAmpRate_thres4 %>%
  filter(grepl('_01_', Chrom))%>% ggplot(aes(x = Pos, y = AmpRate)) + 
  geom_point(alpha = .3, size = .25) +
  theme_bw()+
  theme(axis.text.x = element_blank())+
    labs(y = 'Amp. rate (thres = 5)',
         x = 'Venezuela'),
  x = .5,
  y = .25,
  heigh = .25,
  width = .5
  ) + 
  draw_plot(
    Read_Depth_Summ %>%
  filter(grepl('_01_', Chrom), Country == 'Peru')%>% ggplot(aes(x = Pos, y = mean_ReadDepth)) + 
  geom_line() +
  facet_grid(.~Chrom)+
  theme_bw()+
  theme(axis.text.x = element_blank(),
        legend.position = 'none',
        axis.title.x = element_blank())+
    labs(y = 'Mean read depth'),
  x = 0,
  y = .75,
  heigh = .25,
  width = .5
  )+
  draw_plot(
    Peru_LocusAmpRate_thres0 %>%
  filter(grepl('_01_', Chrom))%>% ggplot(aes(x = Pos, y = AmpRate)) + 
  geom_point(alpha = .3, size = .25) +
  theme_bw()+
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank())+
    labs(y = 'Amp. rate (thres = 1)'),
  x = 0,
  y = .5,
  heigh = .25,
  width = .5
  )+
  draw_plot(
    Peru_LocusAmpRate_thres4 %>%
  filter(grepl('_01_', Chrom))%>% ggplot(aes(x = Pos, y = AmpRate)) + 
  geom_point(alpha = .3, size = .25) +
  theme_bw()+
  theme(axis.text.x = element_blank())+
    labs(y = 'Amp. rate (thres = 5)',
         x = 'Peru'),
  x = 0,
  y = .25,
  heigh = .25,
  width = .5
  ) +
  draw_plot(data.frame(Country = 'Venezuela',
           mean_ReadDepth = Read_Depth_Summ %>%
  filter(grepl('_01_', Chrom), Country == 'Venezuela') %>%
  select(mean_ReadDepth) %>% unlist(),
  LocusAmpRate = Venezuela_LocusAmpRate_thres4 %>%
  filter(grepl('_01_', Chrom)) %>%
  select(AmpRate) %>% unlist()
  )%>%
  ggplot(aes(x = mean_ReadDepth, y = LocusAmpRate))+
  geom_point(alpha = .3)+
  theme_bw()+
  theme(legend.position = 'none') + 
    labs(y = 'Amp. rate (thres = 5)',
         x = 'Mean RD'),
  x = .5,
  y = 0,
  heigh = .25,
  width = .5) +
  draw_plot(data.frame(Country = 'Peru',
           mean_ReadDepth = Read_Depth_Summ %>%
  filter(grepl('_01_', Chrom), Country == 'Peru') %>%
  select(mean_ReadDepth) %>% unlist(),
  LocusAmpRate = Venezuela_LocusAmpRate_thres4 %>%
  filter(grepl('_01_', Chrom)) %>%
  select(AmpRate) %>% unlist()
  )%>%
  ggplot(aes(x = mean_ReadDepth, y = LocusAmpRate))+
  geom_point(alpha = .3)+
  theme_bw()+
  theme(legend.position = 'none') + 
    labs(y = 'Amp. rate (thres = 5)',
         x = 'Mean RD'),
  x = 0,
  y = 0,
  heigh = .25,
  width = .5)
  
```

11. Mask alleles with less than 5 reads of coverage 

```{r}
# rm(Peru_rGenome)
# rm(Venezuela_rGenome)
Pviv_rGenome_thres5 = Pviv_rGenome_object

Pviv_rGenome_thres5@gt = prune_alleles(Pviv_rGenome_thres5, threshold = 4, n = 100)
```

12. Check the distribution of the amplification rate of the samples using a histogram

```{r}

Pviv_rGenome_thres5@metadata$SampAmpRate = SampAmpRate_summ %>% filter(Threshold == 5) %>% select(AmpRate) %>% unlist

Pviv_rGenome_thres5@metadata %>% ggplot(aes(x = SampAmpRate, fill = Country))+
  geom_histogram(position = 'stack', binwidth = .05, color = 'gray40')+
  theme_bw()+
  labs(title = 'Sample Amplification Rate distribution',
       x = 'Amplification Rate')

```

13 Remove samples with less than 50% of amplified loci

```{r}
Pviv_rGenome_thres5 = filter_samples(obj = Pviv_rGenome_thres5, v =
                Pviv_rGenome_thres5@metadata$SampAmpRate >= .5)
```

14. Remove monomorphic sites

14.1 Update allele counts

```{r}

allele_counts = NULL

test = get_AC(obj = Pviv_rGenome_thres5, w = w, n = 100)

for(w in 1:100){
  allele_counts = rbind(allele_counts, get_AC(obj = Pviv_rGenome_thres5, w = w, n = 100))
  print(w)
}

Pviv_rGenome_thres5@loci_table = cbind(Pviv_rGenome_thres5@loci_table,
                                       allele_counts)
```

14.2 Filter monomorphic sites
```{r}
Pviv_rGenome_thres5 = filter_loci(Pviv_rGenome_thres5,
                             v = Pviv_rGenome_thres5@loci_table$Cardinality > 1)
```

15. Remove loci with high missing data

15.1 Calculate the amplification rate of each loci
```{r}
Pviv_rGenome_thres5 = LocusAmplRate(Pviv_rGenome_thres5)
```

15.2 Check the distribution of the amplification rate of the samples using a histogram

```{r}
Pviv_rGenome_thres5@loci_table %>%
    mutate(CHROM2  = gsub('(^(\\d|P|v)+_|_v1)', '',CHROM))%>%
    ggplot(aes(x = POS, y = LocusAmplRate)) +
    geom_point(alpha = 0.7, size = .25) +
    facet_grid(.~ CHROM2)+
    theme_bw()+
    labs(y = 'Ampl. Rate', x = 'Chromosomal position')+
    theme(legend.position = 'right',
          axis.text.x = element_blank())

```

15.3 Filter loci with amplification rate below .75
```{r}
Pviv_rGenome_thres5 = filter_loci(Pviv_rGenome_thres5, v = Pviv_rGenome_thres5@loci_table$LocusAmplRate >= .75)
```

16. Update alternative alleles

```{r}
Pviv_rGenome_thres5@loci_table$ALT =
  ifelse(Pviv_rGenome_thres5@loci_table$REF !=
           gsub(',([ATCG]|\\*)+',
                '',
                gsub(':\\d+',
                     '',
                     Pviv_rGenome_thres5@loci_table$Alleles)),
         gsub(':\\d+',
              '',
              Pviv_rGenome_thres5@loci_table$Alleles),
         gsub('^([ATCG]|\\*)+,',
              '',
              gsub(':\\d+', '', Pviv_rGenome_thres5@loci_table$Alleles))
       )
```

### Step 3: Filter of potential PCR genotyping artifacts (Homopolymers & Short Tandem Repeats)

#### Step 3.1 Clasify SNPs, INDELS, Homopolymers and Short tandem repeats
17. Differentiate between SNPs, INDELS, Homopolymers and Short tandem repeats

```{r}
Pviv_rGenome_thres5@loci_table$TypeOf_Markers =
  TypeOf_Marker(Pviv_rGenome_thres5, w = 1, n = 1)
```

#### Step 3.2 Effect of PCR genotyping artifacts on heterozygosity

18. Get Proportion of Heterozygous samples per site

```{r}
Pviv_rGenome_thres5@loci_table$ObsHet = get_ObsHet(Pviv_rGenome_thres5, by = 'loci', w = 1, n = 1)
```

19. Calculate fraction of Heterozygous samples per Alternative alleles per site

```{r}
Pviv_rGenome_thres5@loci_table$frac_ofHet_pAlts =
  frac_ofHet_pAlt(Pviv_rGenome_thres5, w = 1, n = 1)
```

20. Classify sites based on the proportion of alternative alleles in polyclonal samples per site

20.1 Distribution of the fraction of alternative alleles in polyclonal samples

```{r}
Pviv_rGenome_thres5@loci_table %>% ggplot(aes(x = frac_ofHet_pAlts))+
    geom_histogram(binwidth = 0.01)+
    labs(x = 'Fraction of heterozygous samples per alternative allele per site',
         y = 'Number of sites (Loci)')+
    theme_bw()
```
20.2 Classify sites based on the proportion of alternative alleles in polyclonal samples per site
```{r}
Pviv_rGenome_thres5@loci_table %<>% mutate(ALT_FILTER =case_when(
    frac_ofHet_pAlts == 1 ~ '100%',
    frac_ofHet_pAlts < 1 & frac_ofHet_pAlts > .5 ~ '50 - 99%',
    frac_ofHet_pAlts <= .5 ~ '<=50%',
  ))
```

20.3 Distribution of observed heterozygosity per locus, by type of marker, by group (defined in previuos step 10.2)
```{r}
Pviv_rGenome_thres5@loci_table %>%
    ggplot(aes(x = ObsHet,
               fill = factor(ALT_FILTER,
                             levels = c('<=50%', '50 - 99%', '100%'))))+
    geom_histogram(binwidth = .01)+
    scale_fill_manual(values = c('dodgerblue3', 'gold3', 'firebrick3'))+
    facet_wrap(.~factor(TypeOf_Markers, levels = c(
      'SNP',
      'INDEL',
      'INDEL:Homopolymer',
      'INDEL:Dinucleotide_STR',
      'INDEL:Trinucleotide_STR',
      'INDEL:Tetranucleotide_STR',
      'INDEL:Pentanucleotide_STR',
      'INDEL:Hexanucleotide_STR'
    )), scales = 'free_y', ncol = 4)+
    labs(y = 'Number of Loci',
         x = 'Observed Heterozygosity per locus',
         fill = 'Het/Alt')+
    theme_bw()
```

21. Remove Homopolymers and DiSTRs

```{r}
Pviv_rGenome_thres5 = filter_loci(Pviv_rGenome_thres5,
            v = !(Pviv_rGenome_thres5@loci_table$TypeOf_Markers %in%
                    c('INDEL:Homopolymer', 'INDEL:Dinucleotide_STR')))
```

### Step 4: Filter of problematic genomic regions to map the reads

#### Step 4.1: Regions with excess of Heterozygosity
22. Distribution of proportion of heterozygous per site by sites with \<= 50% Het/Alt, 50 - 99% Het/Alt and 100% of Het/Alt

```{r}
  # In that way we identify that maximum fraction of Heterozygous samples per loci is 0.4855491
  
Pviv_rGenome_thres5@loci_table %>%
    mutate(TypeOf_Markers2 = case_when(
      TypeOf_Markers == 'SNP' ~ 'SNP',
      TypeOf_Markers != 'SNP' ~ 'INDELs'),
      CHROM2  = gsub('(^(\\d|P|v)+_|_v1)', '',CHROM))%>%
    ggplot(aes(x = POS, y = ObsHet, color = factor(ALT_FILTER, levels = c('<=50%', '50 - 99%', '100%')))) +
    geom_point(alpha = 0.5, size = .25) +
    scale_color_manual(values = c('dodgerblue3', 'gold3', 'firebrick3'))+
    facet_grid(TypeOf_Markers2 ~ CHROM2)+
    theme_bw()+
    labs(y = 'Observed Heterozygosity', x = 'Chromosomal position', color = 'Het/Alt')+
    theme(legend.position = 'right',
          axis.text.x = element_blank(),
          axis.title.x = element_blank())
```

23. Calculate the average of heterozygous samples per site per gene

```{r}
Pviv_rGenome_thres5@loci_table = mean_ObsHet(Pviv_rGenome_thres5, gff = '../reference/genes.gff')
```

24. Define a threshold for trusted variants
```{r}
ObsHet_Threshold = quantile(Pviv_rGenome_thres5@loci_table %>%
                                filter(ALT_FILTER == '<=50%', TypeOf_Markers == 'SNP') %>%
                                group_by(gene_id) %>%
                                dplyr::summarise(mean_ObsHet = max(mean_ObsHet)) %>%
                                select(mean_ObsHet) %>%
                                unlist, .95)
```

25. Visualize the Average Observed Heteozygosity in SNPs per genomic position
```{r}
Pviv_rGenome_thres5@loci_table %>%
    filter(ALT_FILTER == '<=50%', TypeOf_Markers == 'SNP') %>%
    group_by(gene_id) %>%
    dplyr::summarise(mean_ObsHet = max(mean_ObsHet)) %>%
    ggplot(aes(x = mean_ObsHet))+
    geom_histogram(binwidth = .005)+
    geom_vline(xintercept = ObsHet_Threshold)+
    labs(y = 'Number of genomic regions', x = 'Average Observed Heteozygosity in SNPs') +
    theme_bw()
```

26. Differentiate genomic regions that PASS the threshold
```{r}
Pviv_rGenome_thres5@loci_table %<>% mutate(ObsHet_Filter = case_when(
    mean_ObsHet > ObsHet_Threshold ~ 'OUT',
    mean_ObsHet <= ObsHet_Threshold ~ 'PASS'
  ))
```

27. Visualize which genomic regions PASS the filter
```{r}
Pviv_rGenome_thres5@loci_table %>%
    mutate(TypeOf_Markers2 = case_when(
      TypeOf_Markers == 'SNP' ~ 'SNP',
      TypeOf_Markers != 'SNP' ~ 'INDEL'),
      CHROM2  = gsub('(^(\\d|P|v)+_|_v1)', '',CHROM))%>%
    ggplot(aes(x = POS, y = ObsHet, color = ObsHet_Filter)) +
    geom_point(alpha = 0.5, size = .25) +
    geom_hline(yintercept = ObsHet_Threshold)+
    scale_color_manual(values = c('firebrick2', 'dodgerblue3'))+
    facet_grid(TypeOf_Markers2 ~ CHROM2)+
    theme_bw()+
    labs(y = 'Observed Heterozygosity',
         x = 'Chromosomal position',
         color = 'Mean ObsHet > Th')+
    theme(legend.position = 'right',
          axis.text.x = element_blank())
```

#### Step 4.2: Regions with excess of density of Polymorphism

28. Calculate the density of SNPs per genomic region and define a threshold
```{r}
Pviv_rGenome_thres5@loci_table = SNP_density(Pviv_rGenome_thres5, gff = '../reference/genes.gff')

SNP_density_threshold = quantile(Pviv_rGenome_thres5@loci_table %>%
                                     filter(TypeOf_Markers == 'SNP') %>%
                                     group_by(gene_id) %>%
                                     dplyr::summarise(SNP_density = max(SNP_density)) %>%
                                     select(SNP_density) %>%
                                     unlist, .925)
```

29. Distribution of SNP density
```{r}
Pviv_rGenome_thres5@loci_table %>%
    filter(TypeOf_Markers == 'SNP') %>%
    group_by(gene_id) %>%
    dplyr::summarise(SNP_density = max(SNP_density)) %>%
    ggplot(aes(x = SNP_density))+
    geom_histogram(binwidth = .001)+
    geom_vline(xintercept = SNP_density_threshold)+
    labs(y = 'Number of genomic regions', x = 'SNP density') +
    theme_bw()
```

30. Differentiate/Filter genomic regions that PASS the threshold
```{r}
Pviv_rGenome_thres5@loci_table %<>% 
  mutate(SNP_density_Filter = case_when(
    SNP_density > SNP_density_threshold ~ 'OUT',
    SNP_density <= SNP_density_threshold ~ 'PASS'
  ))
```

31. Visualize which genomic regions PASS the filter
```{r}
Pviv_rGenome_thres5@loci_table %>%
    mutate(TypeOf_Markers2 = case_when(
      TypeOf_Markers == 'SNP' ~ 'SNP',
      TypeOf_Markers != 'SNP' ~ 'INDEL'),
      CHROM2  = gsub('(^(\\d|P|v)+_|_v1)', '',CHROM))%>%
    ggplot(aes(x = POS, y = ObsHet, color = SNP_density_Filter)) +
    geom_point(alpha = 0.5, size = .25) +
    scale_color_manual(values = c('firebrick2', 'dodgerblue3'))+
    facet_grid(TypeOf_Markers2 ~ CHROM2)+
    theme_bw()+
    labs(y = 'Observed Heterozygosity', x = 'Chromosomal position', color = 'SNP density')+
    theme(legend.position = 'right',
          axis.text.x = element_blank())
```

32. Differentiate of problematic genomic regions to map reads

```{r}
Pviv_rGenome_thres5@loci_table %<>%
  mutate(Alignment_Filter = case_when(
    ObsHet_Filter == 'OUT' & SNP_density_Filter == 'OUT' ~ 'OUT',
    !(ObsHet_Filter == 'OUT' & SNP_density_Filter == 'OUT') ~ 'PASS'
  ))
```

33. Visualize which genomic regions PASS the filter
```{r}
Pviv_rGenome_thres5@loci_table %>%
    mutate(TypeOf_Markers2 = case_when(
      TypeOf_Markers == 'SNP' ~ 'SNP',
      TypeOf_Markers != 'SNP' ~ 'INDEL'),
      CHROM2  = gsub('(^(\\d|P|v)+_|_v1)', '',CHROM))%>%
    ggplot(aes(x = POS, y = ObsHet, color = Alignment_Filter)) +
    geom_point(alpha = 0.5, size = .25) +
    scale_color_manual(values = c('firebrick2', 'dodgerblue3'))+
    facet_grid(TypeOf_Markers2 ~ CHROM2)+
    theme_bw()+
    labs(y = 'Observed Heterozygosity', x = 'Chromosomal position', color = 'Alignment\nFilter')+
    theme(legend.position = 'right',
          axis.text.x = element_blank())

```

34. Check which genomic regions or genes are removed
```{r}
Pviv_rGenome_thres5@loci_table %>% 
  filter(Alignment_Filter == 'OUT') %>% 
  select(gene_id) %>% 
  unique()
```

35. Remove problematic genomic regions to map reads
```{r}
Pviv_rGenome_filtered = 
  filter_loci(Pviv_rGenome_thres5,
              v = Pviv_rGenome_thres5@loci_table$Alignment_Filter == 'PASS')
```

# Section 2: Merging with other datasets (Pending)

```{r}
write.table(Pviv_rGenome_filtered@loci_table %>% select(CHROM, POS), 'selected_pos.bed', row.names = FALSE, quote = FALSE)

load('genomics_Pvivax_pv4.RData')
load('genomics_Pvivax_LAC.RData')

# Merging with Colombian data set----

# Add metadata
Colombia_rGenome@metadata = merge(Colombia_rGenome@metadata, external_metadata, by = 'Sample_id', all.x = T)

## Remove controls
Colombia_rGenome = filter_samples(Colombia_rGenome, v = grepl('^(SP|Col)',Colombia_rGenome@metadata$Sample_id))

## Calculate the Amplification rate of the samples and remove samples with Amplification rate bellow 50%
Colombia_rGenome = SampleAmplRate(Colombia_rGenome, update = T)

Colombia_rGenome = filter_samples(Colombia_rGenome, v = Colombia_rGenome@metadata$SampleAmplRate >= .5)

## Update alleles
Colombia_rGenome@loci_table = cbind(Colombia_rGenome@loci_table, get_AC(Colombia_rGenome, w = 1, n = 1, update_alleles = T))

## Calculate the amplification rate of each locus
Colombia_rGenome = LocusAmplRate(Colombia_rGenome, update = T)

## Remove loci with amplification rate bellow .75
Colombia_rGenome = filter_loci(Colombia_rGenome, v = Colombia_rGenome@loci_table$LocusAmplRate >= .75)

# Merge PerVen_rGenome with Colombia_rGenome

PerVenCol_rGenome = merge_rGenome(Pviv_rGenome_filtered, Colombia_rGenome)

PerVenCol_rGenome@loci_table = cbind(PerVenCol_rGenome@loci_table, get_AC(PerVenCol_rGenome))

PerVenCol_rGenome@loci_table = PerVenCol_rGenome@loci_table[-5]

PerVenCol_rGenome = filter_samples(PerVenCol_rGenome, v = !grepl('^NA', rownames(PerVenCol_rGenome@metadata)))

save.image('backup.RData')

# Merging with Guyana data set----

## Remove data already present in PerVenCol_rGenome

Guyana_rGenome = filter_samples(Guyana_rGenome, v = !(Guyana_rGenome@metadata$Sample_id %in% PerVenCol_rGenome@metadata$Sample_id))

## Add Metadata

Guyana_rGenome@metadata = data.frame(Sample_id = Guyana_rGenome@metadata)

Guyana_rGenome@metadata %<>% mutate(Country = case_when(
  grepl('^G', Sample_id) ~ 'Guyana',
  grepl('^Nicanor', Sample_id) ~ 'Panama?'
))

rownames(Guyana_rGenome@metadata) = Guyana_rGenome@metadata$Sample_id

Guyana_rGenome = filter_samples(Guyana_rGenome, v = !is.na(Guyana_rGenome@metadata$Country))

# Calculate the Amplification rate of the samples and remove samples with Amplification rate bellow 50%
Guyana_rGenome = SampleAmplRate(Guyana_rGenome, update = T)

Guyana_rGenome = filter_samples(Guyana_rGenome, v = Guyana_rGenome@metadata$SampleAmplRate >= .5)

# Update alleles
Guyana_rGenome@loci_table = cbind(Guyana_rGenome@loci_table, get_AC(Guyana_rGenome, w = 1, n = 1, update_alleles = T))

# Calculate the amplification rate of each locus
Guyana_rGenome = LocusAmplRate(Guyana_rGenome, update = T)

# Remove loci with amplification rate bellow .75
Guyana_rGenome = filter_loci(Guyana_rGenome, v = Guyana_rGenome@loci_table$LocusAmplRate >= .75)

# Merge PerVen_rGenome with Guyana_rGenome

View(Guyana_rGenome@loci_table)

PerVenColGuy_rGenome = merge_rGenome(PerVenCol_rGenome, Guyana_rGenome)

PerVenColGuy_rGenome@loci_table = cbind(PerVenColGuy_rGenome@loci_table, get_AC(PerVenColGuy_rGenome))

PerVenColGuy_rGenome@loci_table = PerVenColGuy_rGenome@loci_table[-5]

names(PerVenColGuy_rGenome@loci_table)

save.image('backup.RData')



```





# Section 3: Differentiating monoclonal and polyclonal infections

36. Define the fraction of heterozygous loci per sample

```{r}
Pviv_rGenome_filtered@metadata$fracHet = 
  get_ObsHet(obj = Pviv_rGenome_filtered, by = 'sample')

```

37. Within host divergence (Fws) per sample

```{r}
Pviv_rGenome_filtered@metadata$Fws = 
  get_Fws(obj = Pviv_rGenome_filtered, w = 1, n = 1)
```

38. Defining Monoclonal, Highly related polyclonal, and Polyclonal infections

```{r}
Pviv_rGenome_filtered@metadata %<>% 
  mutate(Clonality = case_when(
    Fws >= .95 & fracHet <= .015 ~ 'Monoclonal',
    !((Fws >= .95 & fracHet <= .015) | (Fws < .92 | fracHet > .03)) ~ 'Highly related Polyclonal',
    Fws < .92 | fracHet > .03 ~ 'Polyclonal'
  ))
```

39. Visualize distribution of Heterozygosity and Fws
```{r}
Pviv_rGenome_filtered@metadata %>%
  ggplot(aes(x = fracHet, y = Fws, color = Clonality))+
  geom_point()+
  geom_hline(yintercept = .95)+
  geom_hline(yintercept = .92)+
  geom_vline(xintercept = .015)+
  geom_vline(xintercept = .03)+
  facet_grid(.~Country)+
  labs(x = 'Fraction of heterozygous loci')+
  theme_bw()
```

40. Save data
```{r}
rm(list = ls()[-grep('Pviv_rGenome_filtered',ls())])

save.image('../GeneticDiversity/PerVen_Pviv_filtered_rGenome.RData')
```
